---
title: "Preliminary analysis of Wadden Sea sample W8"
author: Pieter Provoost
output: 
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

The goal of this notebook is to do an in-depth analysis of a single sample in order to improve [our eDNA analysis pipeline](https://github.com/iobis/PacMAN-pipeline). The compilation of the reference databases used is documented [here](https://github.com/iobis/eDNA_trial_data/blob/master/eDNA_expeditions_ref_dbs.md).

<style>
pre {
  background-color: '#eeeeee'
}
.tocify-extend-page {
  height: 0 !important;
}
</style>

## Dependencies

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(chameleon)
library(viridis)
library(ggplot2)
library(ggvenn)
library(tidyr)
library(reactablefmtr)
```

## Load results for single sample

This loads all pipeline results from the `runs` folder which is copied from the PacMAN pipeline.

```{r warning=FALSE}
w8_ncbi <- read.csv("runs/dev_coi_ncbi_w8/05-dwca/Occurence_table.csv", sep = "\t", na.strings = c("", "nan")) %>%
  mutate(otu_db = "NCBI", marker = "COI")
w8_16s <- read.csv("runs/dev_16s_w8/05-dwca/Occurence_table.csv", sep = "\t", na.strings = c("", "nan")) %>%
  mutate(otu_db = "NCBI", marker = "16S")
w8_mifish <- read.csv("runs/dev_mifish_w8/05-dwca/Occurence_table.csv", sep = "\t", na.strings = c("", "nan")) %>%
  mutate(otu_db = "NCBI", marker = "MiFish")
w8_mimammal <- read.csv("runs/dev_mimammal_w8/05-dwca/Occurence_table.csv", sep = "\t", na.strings = c("", "nan")) %>%
  mutate(otu_db = "NCBI", marker = "MiMammal")
w8_teleo<- read.csv("runs/dev_teleo_w8/05-dwca/Occurence_table.csv", sep = "\t", na.strings = c("", "nan")) %>%
  mutate(otu_db = "NCBI", marker = "Teleo")

w8_combined <- bind_rows(w8_ncbi, w8_16s, w8_mifish, w8_mimammal, w8_teleo)
```

## Comparison PCR approaches and markers

This compares results by PCR approach (singleplex versus multiplex) and marker.

### Reads

```{r warning=FALSE, out.width="100%", message=FALSE, fig.width=12, fig.height=8}
stats_reads <- w8_combined %>%
  group_by(phylum, eventID, marker) %>%
  summarize(reads = sum(organismQuantity)) %>%
  ungroup() %>%
  mutate(identified = factor(!is.na(phylum), levels = c(FALSE, TRUE)))

ggplot(data = stats_reads) +
  geom_bar(aes(y = phylum, x = reads, fill = marker, alpha = identified), stat = "identity") +
  scale_x_continuous(trans = "log10") +
  scale_fill_viridis(discrete = TRUE, option = "magma", begin = 0.2, end = 0.8) +
  scale_alpha_manual(values = c(0.3, 1)) +
  facet_grid(eventID~marker) +
  ggtitle("Reads by PCR approach and marker")
```

### ASVs

```{r warning=FALSE, out.width="100%", message=FALSE, fig.width=12, fig.height=8}
stats_asvs <- w8_combined %>%
  group_by(phylum, eventID, marker) %>%
  summarize(asvs = n()) %>%
  ungroup() %>%
  mutate(identified = factor(!is.na(phylum), levels = c(FALSE, TRUE)))

ggplot(data = stats_asvs) +
  geom_bar(aes(y = phylum, x = asvs, fill = marker, alpha = identified), stat = "identity") +
  scale_x_continuous(trans = "log10") +
  scale_fill_viridis(discrete = TRUE, option = "magma", begin = 0.2, end = 0.8) +
  scale_alpha_manual(values = c(0.3, 1)) +
  facet_grid(eventID~marker) +
  ggtitle("ASVs by PCR approach and marker")
```

### Species

```{r warning=FALSE, out.width="100%", message=FALSE, fig.width=12, fig.height=8}
stats_species <- w8_combined %>%
  group_by(phylum, eventID, marker) %>%
  filter(taxonRank == "species") %>%
  summarize(species = length(unique(scientificName))) %>%
  ungroup() %>%
  mutate(identified = factor(!is.na(phylum), levels = c(FALSE, TRUE)))

ggplot(data = stats_species) +
  geom_bar(aes(y = phylum, x = species, fill = marker, alpha = identified), stat = "identity") +
  scale_x_continuous() +
  scale_fill_viridis(discrete = TRUE, option = "magma", begin = 0.2, end = 0.8) +
  scale_alpha_manual(values = c(0.3, 1), drop = FALSE) +
  facet_grid(eventID~marker) +
  ggtitle("Species by PCR approach and marker")
```

### Species overlap between PCR approaches

```{r warning=FALSE}
approaches <- unique(w8_combined$eventID)

species_by_approach <- lapply(approaches, function(approach) {
  w8_combined %>%
    filter(eventID == approach & taxonRank == "species") %>% pull(scientificName) %>% unique()
}) %>% setNames(approaches)

ggvenn(
  species_by_approach, 
  fill_color = c("#0073C2FF", "#EFC000FF", "#868686FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 4
)
```

### Species overlap between markers

```{r warning=FALSE}
markers <- c("COI", "MiFish/MiMammal", "16S", "Teleo")

species_by_marker <- lapply(markers, function(m) {
  w8_combined %>%
    mutate(marker = ifelse(marker %in% c("MiFish", "MiMammal"), "MiFish/MiMammal", marker)) %>%
    filter(marker == m & taxonRank == "species") %>% pull(scientificName) %>% unique()
}) %>% setNames(markers)

ggvenn(
  species_by_marker, 
  fill_color = c("#0073C2FF", "#EFC000FF", "#868686FF", "#CD534CFF"),
  stroke_size = 0.5, set_name_size = 4
)
```

## Species lists

### Top species

```{r warning=FALSE, out.width="100%", message=FALSE, fig.width=12, fig.height=10}
top_species <- w8_combined %>%
  filter(taxonRank == "species") %>%
  group_by(phylum, scientificName) %>%
  summarize(asvs = n(), reads = sum(organismQuantity)) %>%
  arrange(desc(reads)) %>%
  head(30) %>%
  arrange(phylum) %>%
  pull(scientificName)

top_species_stats <- w8_combined %>%
  filter(taxonRank == "species" & scientificName %in% top_species) %>%
  group_by(phylum, scientificName, eventID, marker) %>%
  summarize(asvs = n(), reads = sum(organismQuantity)) %>%
  mutate(scientificName = factor(scientificName, levels = top_species))

ggplot(top_species_stats) +
  geom_bar(aes(x = reads, y = scientificName, fill = phylum), stat = "identity") +
  facet_grid(eventID ~ marker) +
  scale_x_continuous(trans = "log10") +
  ggtitle("W8 - top species") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#ffffff", color = "#000000"),
    strip.background = element_blank()
  ) +
  scale_y_discrete(expand = c(0.05, 0.05)) +
  scale_fill_brewer(palette = "Paired", na.value = "#eeeeee")
```

### All species

```{r warning=FALSE, message=FALSE}
species_pcr <- w8_combined %>%
  filter(taxonRank == "species") %>%
  group_by(scientificName) %>%
  summarize(
    single = any(eventID == "W-8-singleplex"),
    multi = any(eventID == "W-8-multi-tot")
  )

all_species <- w8_combined %>%
  filter(taxonRank == "species") %>%
  group_by(scientificName, phylum, marker) %>%
  summarize(reads = sum(organismQuantity)) %>%
  spread(marker, reads) %>%
  left_join(species_pcr, by = "scientificName")

color_scale <- function(x) rgb(colorRamp(c("#ffffff", "#f2fbd2", "#c9ecb4", "#93d3ab", "#35b0ab"))(x), maxColorValue = 255)

number_def <- colDef(
  maxWidth = 75,
  style = function(value) {
    if (is.numeric(value) & !is.na(value)) {
      normalized <- log10(value + 1) / 8
      color <- color_scale(normalized)
      list(background = color, "font-size" = "0.9em")
    }
  }
)

boolean_def <- colDef(
  maxWidth = 50,
  style = function(value) {
    if (value == TRUE) {
      list(background = "#FFF3E0", "font-size" = "0.9em")
    } else {
      list()
    }
  }
)

text_def <- colDef(
  style = list("font-size" = "0.95em")
)

reactable(all_species, defaultSorted = "phylum", pagination = FALSE, highlight = TRUE, compact = TRUE, wrap = FALSE, columns = list(
  "scientificName" = text_def,
  "phylum" = text_def,
  "single" = boolean_def,
  "multi" = boolean_def,
  "16S" = number_def,
  "COI" = number_def,
  "MiFish" = number_def,
  "MiMammal" = number_def,
  "Teleo" = number_def
))
```

## Questionable annotations (work in progress)

Some of the annotations are questionable and need to be looked into. For example: Aurelia miyakei, Argentina kagoshimae.
